### 2016-10-03 Paper_v7.tex
First pass through by JV

y\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses
\usepackage{ifthen}  % Conditional
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{lineno}

@@

\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Population heterogeneity in the epithelial to mesenchymal transition is controlled by NFAT and phosphorylated Sp1}}
\vspace{0.05in}
{\par \centering Russell Gould$^{\ddag}$ and David M. Bassen, Anirikh Chakrabarti, Jeffrey D. Varner and Jonathan Butcher$^{\ddag *}$}
\vspace{0.05in}
{\par \centering School of Chemical and Biomolecular Engineering and $^{\ddag}$Department of Biomedical Engineering, Cornell University, Ithaca NY 14853}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~ Modeling of TGF-$\beta$ induced EMT}
\vspace{0.1in}
% {\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology (60,000 Character Limit)}}
%\vspace{0.1in}
{\par \centering \textbf{Character count:}~\emph{56,659}}
\vspace{0.1in}
{\par \centering $^{*}$Corresponding author:}
\vspace{0.1in}
{\par \centering Jonathan T. Butcher,}
{\par \centering Associate Professor, Department of Biomedical Engineering,}
{\par \centering 304 Weill Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jtb47@cornell.edu}
{\par \centering Phone: (607) 255 - 3575}
{\par \centering Fax: (607) 255 - 7330}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
\section*{Abstract}
Epithelial to mesenchymal transition (EMT) is an essential differentiation program during tissue morphogenesis and remodeling.
EMT is induced by soluble transforming growth factor $\beta$ (TGF-$\beta$) family members, and restricted by vascular endothelial growth factor family members.
While many downstream molecular regulators of EMT have been identified, these have been largely evaluated individually without considering potential crosstalk.
In this study, we created an ensemble of dynamic mathematical models describing TGF-$\beta$ induced EMT to better understand the operational hierarchy of this complex molecular program.
These models incorporate mass action kinetics within an ordinary differential equation (ODE) framework to describe the transcriptional and post-translational regulatory events driving EMT.
Model parameters were estimated from multiple data sets using multiobjective optimization, in combination with cross-validation.
TGF-$\beta$ exposure drove the model population toward a mesenchymal phenotype, while an epithelial phenotype was maintained following vascular endothelial growth factor A (VEGF-A) exposure.
Simulations predicted that the transcription factors phosphorylated SP1 and NFAT were master regulators promoting or inhibiting EMT, respectively.
Surprisingly, simulations also predicted that a cellular population could exhibit phenotypic heterogeneity (characterized by a significant fraction of the population with both high epithelial and mesenchymal marker expression)
if treated simultaneously with TGF-$\beta$ and VEGF-A.  We tested this prediction experimentally in both MCF10A and DLD1 cells and found that upwards of 45\% of the cellular population acquired this hybrid state in the presence of both TGF-$\beta$ and VEGF-A.
We experimentally validated the predicted NFAT/Sp1 signaling axis for each phenotype response. Lastly, we found that cells in the hybrid state had significantly different functional behavior when compared to VEGF-A or TGF-$\beta$ treatment alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and potentially reveal a novel signaling axis which regulates carcinoma progression through an EMT versus tubulogenesis response.

@@

In this study, we developed a family of mechanistic models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of vascular endothelial growth factor A (VEGF-A).
We incorporated mass action kinetics within an ordinary differential equation (ODE) framework to describe the EMT interaction network containing 97 gene, protein or mRNA components interconnected through 169 interactions.
A family of model parameters was estimated using 41 molecular data sets generated in DLD1 colon carcinoma, MDCKII and A375 melanoma cells using the Pareto optimal ensemble technique (POETs) multiobjective optimization algorithm.
POETs generated an ensemble of approximately 1400 models for analysis.
Analysis of the model population suggested that both MCF10A and DLD1 cells could exhibit phenotypic heterogeneity if treated simultaneously with TGF-$\beta$1/2 and VEGF-A.
This heterogeneity was characterized by a significant fraction of the population being in a “hybrid state” having both high E-cadherin and high Vimentin expression.
We tested these predictions using qRT-PCR and flow-cytometry studies in a variety of experimental conditions.
Validation studies confirmed that upwards of 45\% of the cellular population could be put into the hybrid state in the presence of both TGF-$\beta$1/2 and VEGF-A.
Moreover, this response depended upon both activation of Sp1 by MAPK and NFATc1 transcriptional activity consistent with the predicted molecular signaling.
Lastly, the hybrid populations of both DLD1 and MCF10A cells exhibited different functional behavior than those from either TGF-$\beta$ or VEGF-A treatment.
The extent of ductal branch formation significantly increased with MCF10A cells in the hybrid phenotype, compared with cells treated with VEGF-A alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and reveal a novel signaling axis, which possibly regulates carcinoma progression through an EMT versus tubulogenesis response.

@@

\subsection*{The model population captured key features of TGF-$\beta$ induced EMT}
The EMT model architecture, based upon curated molecular connectivity,
described the expression of 23 genes following exposure to TGF-$\beta$ isoforms and VEGF-A (Fig. \ref{fg:F1}).
The EMT model contained 74 molecular species interconnected by 169 interactions.
Model equations were formulated using mass-action kinetics within an ordinary differential equation (ODE) framework.
ODEs and mass action kinetics are common tools to model biochemical pathways \citep{Schoeberl:2002rv,Chen:2009uq,Tasseff:2011ve}.
However, while ODE models can simulate complex intracellular behavior,
they require estimates for model parameters which are often difficult to obtain.
The EMT model had 251 unknown model parameters, 169 kinetic constants and 45 non-zero initial conditions.
As expected, these parameters were not uniquely identifiable given the training data \citep{Gadkar:2005vn}.
Thus, instead of identifying a single best fit (but uncertain) model,
we estimated a sloppy population of models (each consistent with the training data) by simultaneously minimizing the difference between model simulations and 41 molecular
data sets using the Pareto Optimal Ensemble Technique (POETs).
The training data were generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells following
exposure to TGF-$\beta$ isoforms \citep{Medici:2008fk}. We organized these data sets into 11 objective functions which were simultaneously minimized by POETs.
Additionally, we used 12 molecular data sets generated in HK-2 cells following VEGF-A exposure to train VEGF-A responsive model processes \citep{Lian:2011bs}.
To guard against overfitting, we augmented the multiobjective optimization with leave-one-out cross validation to independently estimate both the training and
prediction error for each objective. Thus, we generated 11 different model ensembles.
Lastly, we compared model predictions with independent data sets not used during training (both at the molecular and model population
levels) to evaluate the predictive power of the parameter ensemble.
Additional details of the signaling architecture included in the model are presented in the materials and methods and the supplement.

POETs generated a population of probable signaling models which captured the multiple phases of EMT induction (Fig. \ref{fg:F2}).
POETs sampled well over 10$^4$ probable models during each stage of the cross-validation #####, using a combination of both local and global random sampling.
From this analysis, N $\simeq$ 1400 models were selected for further analysis.
The selected models all had the same possible molecular connectivity, but different values for model parameters. Trascription and translation rates, as well as mRNA and protein degradation terms, were set using physical values from the literature, and allowed to vary by a scaling factor, see methods ####citebionumbers####.
Model selection was based upon Pareto rank, the prediction and training error across all objectives and model to model correlation (supplemental materials).
The model population recapitulated key signaling events following TGF-$\beta$ exposure.
We subdivided the response to TGF-$\beta$ exposure into two phases.
First, TGF-$\beta$1/2 signaling initiated a program which downregulated E-cadeherin expression in a MAPK dependent manner while
simultaneously upregulating TGF-$\beta$3 expression. Second, TGF$-\beta$3 secretion initiated an autocrine feedback which upregulated the expression of
mesenchymal markers such as Vimentin and key upstream transcription factors such as LEF-1 in a SMAD dependent manner. TGF-$\beta$3 expression was also able to sustain $/beta$-catenin release by inhibiting its sequestration by the APC complex through PI3K mediated GSK3, which was captured by the model (Fig. \ref{fg:F4B})
Each phase involved the hierarchal expression and/or post-translational modification of several key transcription factors.
During the first phase, stimulation with TGF-$\beta$1/2 (10 a.u.) activated both the SMAD and MAPK pathways.
MAPK activation resulted in the phosphorylation of the transcription factor activator protein 1 (AP-1), which in-turn upregulated the expression of Snail, a well established transcriptional repressor (Fig. \ref{fg:F2}A).
Snail expression was MAPK-dependent; the MEK inhibitor U0126 blocked AP-1 activation and Snail expression following TGF-$\beta$1/2 exposure (Fig. \ref{fg:F2}A, Lane 3).
Similar results were obtained for Slug expression, confirming initial activation through the MAPK pathway (data not shown).
Overexpression of either Snail or Slug upregulated TGF-$\beta$3 expression (Fig. \ref{fg:F2}C) while simultaneously downregulating E-cadeherin expression (Fig. \ref{fg:F2}F).
During the second phase, TGF$-\beta$3 secretion and the subsequent autocrine signaling resulted in the upregulation of mesenchymal marker expression.
The TGF$-\beta$3 induced gene expression program involves a complex hierarchy of transcriptional and post-translational regulatory events.
Absence of E-cadherin indirectly promoted TGF$-\beta$3 expression through the $\beta$-catenin/TCF4 complex following Snail or Slug expression (Fig. \ref{fg:F2}C, Lane 2 or 3).
Conversely, over-expression of E-cadherin inhibited the TGF$-\beta$3 autocrine production by sequestering cytosolic $\beta$-catenin, thereby blocking EMT (Fig. \ref{fg:F2}C, Lane 4 or 5).
TGF$-\beta$3 signaled through the Smad pathway to regulate LEF-1 expression and downstream target EMT genes (Fig. \ref{fg:F2}G).
TGF$-\beta$3 (10 a.u.) in combination with downstream inhibitors (DN-Smad4 and DN-LEF-1) completely inhibited Vimentin expression, while elevating E-cadherin expression (Fig. \ref{fg:F2}H,I).

The predictive power of the ensemble was tested using both cross validation and by comparing simulations with data sets not used for model training.
In whole, 72\% of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
generated from a random starting point.
Conversely, we \emph{predicted} approximately 72\% of the training objectives, at a 95\% confidence interval compared to randomized parameters.
The model also captured the temporal gene expression responses of E-cadherin, pSmad2, and LEF-1 to within one-standard deviation (up to the 48 hr time-point) (Fig. \ref{fg:F2}J-L).
This data was not used for model training.
The high predictability can be attributed to the combination of the leave-one-out cross validation scheme, diverse objective functions, and robustness of the POETs algorithm.
Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \citep{Medici:2008fk} that drive the phenotypic conversion.
A listing of objective function data resulting from training and cross validation is included in the supplement (Fig. \ref{fig:SX}).

\subsection*{Identification of a novel LEF-1 regulator}
During model identification, we found that consistent TGF-$\beta$ induced EMT required an additional regulatory protein.
This protein, which we called hypothetical regulator 1 (YREG1), was required to mediate between SNAIL/SLUG transcriptional activity
and the upregulation of LEF-1 expression following TGF-$\beta$1/2 exposure.
SNAIL/SLUG are well known transcriptional repressors \citep{Hemavathy:2000aa,Hemavathy:2000ab,Dhasarathy:2011aa}, although there are a few studies
which suggest that at least SNAIL can also act as a transcriptional activator \citep{Guaita:2002aa}.
In the model, we assumed the expression of SNAIL/SLUG was likely regulated by AP1/SP1 \citep{Jackstadt:2013aa}.
Thus, upon receiving a TGF-$\beta$1/2 signal, the model predicted enhanced SNAIL/SLUG expression, consistent with experimental observations.
TGF-$\beta$1/2 stimulation also induces LEF-1 expression. However, literature evidence suggested that LEF-1 expression was not strongly dependent upon AP1/SP1 activity \citep{Eastman:1999aa}.
Thus, either SNAIL/SLUG are acting as inducers (contrary to substantial biochemical evidence) or, they are repressing the expression of an intermediate repressor.
Given the biochemical evidence supporting SNAIL/SLUG as repressors, we created YREG1 a hypothetical intermediate repressor whose
expression is downregulated by SNAIL/SLUG. The literature data therefore suggested that YREG1 had two transcriptional targets, LEF-1 and TGF-$\beta$3.
By adding this regulator, our simulations became consistent with training and literature data.
Medici et al. suggested a similar idea where feedback between $\beta$-catenin and LEF-1 was likely, although this feedback had yet to be identified \citep{Medici:2008fk}.
Low levels of YREG1 expression were present in all simulations to regulate the formation of the $\beta$-catenin-LEF-1 complex.
To test the potency of YREG1, we conducted over-expression and knockdown simulations following the addition of TGF-$\beta$1/2 #### (Fig. \ref{fig:4}C and \ref{fig:4}D).
In the absence of YREG1, most of the population failed to consistently respond to TGF-$\beta$1/2 exposure compared to the wild-type (Fig. \ref{fig:S8}A).
Conversely, YREG1 overexpression revealed an exclusively epithelial phenotype following TGF-$\beta$1/2 stimulation (Fig. \ref{fig:S8}B).####
Overexpression of YREG1 repressed LEF-1 and TGF-$\beta$3 expression,
thereby not allowing free $\beta$-catenin to form the $\beta$-catenin-LEF-1 complex which promotes mesenchymal gene expression, or SMAD activity following from autocrine TGF-$\beta$3 signaling.
Taken together, we found that low YREG1 expression was necessary for stabilizing EMT,
while elevated YREG1 levels limited the extent of EMT induction.

@@

However, the most biologically interesting behavior was exhibited by cells achieving a hybrid phenotype, most notable in a dual treatment condition (black arrow), but also present in a small percentage of untreated cells (Fig. \ref{fg:F3}A) .
Models with this hybrid phenotype had elevated Sp1 and NFAT transcriptional activity, resulting in increased \emph{both} Vimentin and E-cadherin expression, respectively (Fig. \ref{fg:F4}A).
Analysis of these hypothetical cells suggested they had \emph{abnormal} signaling; deregulated NFAT expression and nuclear localization promoted E-cadherin expression while TGF$-\beta$1/2 induced Sp1 action promoted Vimentin expression. 
####Analysis of the connectivity and information flow through the signaling architecture suggested that Sp1 and NFAT action could be manipulated \emph{independently} by simultaneous TGF$-\beta$1/2 and VEGF-A stimulation (Fig. \ref{fg:S1}).

To test this hypothesis, we simulated the response of the network to TGF$-\beta$1/2 and VEGF-A treatment with and without NFATc inhibitors (Fig. \ref{fg:F4}).
As expected, stimulation with VEGF-A (50 a.u.) maintained an epithelial population, while TGF$-\beta$1/2 (10 a.u.) exposure shifted
the population from an epithelial to a mesenchymal phenotype (Fig. \ref{fg:F3}A and Fig. \ref{fg:F3}B).
On the other hand, combined stimulation with TGF$-\beta$1/2 (10 a.u.) and VEGF-A (50 a.u.) increased both E-cadherin and Vimentin expression, resulting in a hybrid phenotype with both epithelial and mesenchymal characteristics (Fig. \ref{fg:F3}C).
Vimentin expression was correlated with high levels of nuclear phosphorylated Sp1, following TGF$-\beta$1/2 exposure.
Conversely, elevated E-cadherin expression depended upon the activity of NFAT transcription factors downstream of VEGF-A stimulation.
To further isolate the role of NFAT on this hybrid state, we simulated the inhibition of NFAT transcriptional activity across all conditions (all else being equal).
NFAT inhibition in combination with VEGF-A and TGF$-\beta$1/2 treatments blocks increased activation of E-cadherin (Fig. \ref{fg:3}D and Fig. \ref{fg:3}E).
Lastly, NFATc inhibition in combination with simultaneous TGF$-\beta$1/2 and VEGF-A exposure repressed nearly all E-cadherin expression, shifting nearly the entire population towards a mesenchymal phenotype (Fig. \ref{fg:F4}F). Taken together, high levels of nuclear localized phosphorylated Sp1 correlated with Vimentin expression, while NFATc transcriptional activity was predicted to be critical for maintaining E-cadherin expression in the prescence of competing signals. In particular, when present, VEGF-A signaling through NFATc was predicted to provide a protective up-regulation of E-cadherin.

@@

\section*{Materials and Methods}
The simulation code and parameter ensemble used in this study can be downloaded from GitHub (https://github.com/jeffreyvarner/TGFb-VEGFA-Model\_v1.git).

>>>>

\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses
\usepackage{ifthen}  % Conditional
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{lineno}

@@

\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Population heterogeneity in the epithelial to mesenchymal transition is controlled by NFAT and phosphorylated Sp1}}
\vspace{0.05in}
{\par \centering Russell Gould$^{\ddag}$ and David M. Bassen, Anirikh Chakrabarti, Jeffrey D. Varner and Jonathan Butcher$^{\ddag *}$}
\vspace{0.05in}
{\par \centering Robert Frederick Smith School of Chemical and Biomolecular Engineering and $^{\ddag}$Nancy E. and Peter C. Meinig School of Biomedical Engineering, Cornell University, Ithaca NY 14853}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~ Modeling of TGF-$\beta$ induced EMT}
\vspace{0.1in}
% {\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology (60,000 Character Limit)}}
%\vspace{0.1in}
{\par \centering \textbf{Character count:}~\emph{56,659}}
\vspace{0.1in}
{\par \centering $^{*}$Corresponding author:}
\vspace{0.1in}
{\par \centering Jonathan T. Butcher,}
{\par \centering Associate Professor, Nancy E. and Peter C. Meinig School of Biomedical Engineering,}
{\par \centering 304 Weill Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jtb47@cornell.edu}
{\par \centering Phone: (607) 255 - 3575}
{\par \centering Fax: (607) 255 - 7330}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
\section*{Abstract}
Epithelial to mesenchymal transition (EMT) is an essential differentiation program during tissue morphogenesis and remodeling.
EMT is induced by soluble transforming growth factor $\beta$ (TGF-$\beta$) family members, and restricted by vascular endothelial growth factor family members.
While many downstream molecular regulators of EMT have been identified, these have been largely evaluated individually without considering potential crosstalk.
In this study, we created an ensemble of dynamic mathematical models describing TGF-$\beta$ induced EMT to better understand the operational hierarchy of this complex molecular program.
These models incorporate mass action kinetics within an ordinary differential equation (ODE) framework to describe the transcriptional and post-translational regulatory events driving EMT.
Model parameters were estimated from multiple data sets using multiobjective optimization, in combination with cross-validation.
TGF-$\beta$ exposure drove the model population toward a mesenchymal phenotype, while an epithelial phenotype was maintained following vascular endothelial growth factor A (VEGF-A) exposure.
Simulations predicted that the transcription factors phosphorylated SP1 and NFAT were master regulators promoting or inhibiting EMT, respectively.
Surprisingly, simulations also predicted that a cellular population could exhibit phenotypic heterogeneity (characterized by a significant fraction of the population with both high epithelial and mesenchymal marker expression)
if treated simultaneously with TGF-$\beta$ and VEGF-A.  We tested this prediction experimentally in both MCF10A and DLD1 cells and found that upwards of 45\% of the cellular population acquired this hybrid state in the presence of both TGF-$\beta$ and VEGF-A.
We experimentally validated the predicted NFAT/Sp1 signaling axis for each phenotype response. Lastly, we found that cells in the hybrid state had significantly different functional behavior when compared to VEGF-A or TGF-$\beta$ treatment alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and potentially reveal a novel signaling axis which regulates carcinoma progression through an EMT versus tubulogenesis response.

@@

In this study, we developed a family of mechanistic models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of vascular endothelial growth factor A (VEGF-A).
We incorporated mass action kinetics within an ordinary differential equation (ODE) framework to describe the EMT interaction network containing 97 gene, protein or mRNA components interconnected through 169 interactions.
A family of model parameters was estimated using 41 molecular data sets generated in DLD1 colon carcinoma, MDCKII and A375 melanoma cells using the Pareto optimal ensemble technique (JuPOETs) multiobjective optimization algorithm.
JuPOETs generated an ensemble of approximately 1400 models for analysis.
Analysis of the model population suggested that both MCF10A and DLD1 cells could exhibit phenotypic heterogeneity if treated simultaneously with TGF-$\beta$1/2 and VEGF-A.
This heterogeneity was characterized by a significant fraction of the population being in a “hybrid state” having both high E-cadherin and high Vimentin expression.
We tested these predictions using qRT-PCR and flow-cytometry studies in a variety of experimental conditions.
Validation studies confirmed that upwards of 45\% of the cellular population could be put into the hybrid state in the presence of both TGF-$\beta$1/2 and VEGF-A.
Moreover, this response depended upon both activation of Sp1 by MAPK and NFATc1 transcriptional activity consistent with the predicted molecular signaling.
Lastly, the hybrid populations of both DLD1 and MCF10A cells exhibited different functional behavior than those from either TGF-$\beta$ or VEGF-A treatment.
The extent of ductal branch formation significantly increased with MCF10A cells in the hybrid phenotype, compared with cells treated with VEGF-A alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and reveal a novel signaling axis, which possibly regulates carcinoma progression through an EMT versus tubulogenesis response.

@@

\subsection*{The model population captured key features of TGF-$\beta$ induced EMT}
The EMT model architecture, based upon curated molecular connectivity,
described the expression of 23 genes following exposure to TGF-$\beta$ isoforms and VEGF-A (Fig. \ref{fg:F1}).
The EMT model contained 74 molecular species interconnected by 169 interactions.
Model equations were formulated using either saturation or mass-action kinetics within an ordinary differential equation (ODE) framework.
ODEs and mass action kinetics are common tools to model biochemical pathways \citep{Schoeberl:2002rv,Chen:2009uq,Tasseff:2011ve}.
However, while ODE models can simulate complex intracellular behavior,
they require estimates for model parameters which are often difficult to obtain.
The EMT model had 251 unknown model parameters, 169 kinetic constants and 45 non-zero initial conditions.
As expected, these parameters were not uniquely identifiable given the training data \citep{Gadkar:2005vn}.
Thus, instead of identifying a single best fit (but uncertain) model,
we estimated a sloppy population of models (each consistent with the training data) by simultaneously minimizing the difference between model simulations and 41 molecular
data sets using the Pareto Optimal Ensemble Technique (JuPOETs).
The training data were generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells following
exposure to TGF-$\beta$ isoforms \citep{Medici:2008fk}. We organized these data sets into 11 objective functions which were simultaneously minimized by JuPOETs.
Additionally, we used 12 molecular data sets generated in HK-2 cells following VEGF-A exposure to train VEGF-A responsive model processes \citep{Lian:2011bs}.
To guard against overfitting, we augmented the multiobjective optimization with leave-one-out cross validation to independently estimate both the training and
prediction error for each objective. Thus, we generated 11 different model ensembles.
Lastly, we compared model predictions with independent data sets not used during training (both at the molecular and model population
levels) to evaluate the predictive power of the parameter ensemble.

JuPOETs generated a population of probable signaling models which captured the multiple phases of EMT induction (Fig. \ref{fg:F2}).
JuPOETs sampled well over 10$^4$ probable models during each stage of the cross-validation using global random sampling.
From this analysis, N $\simeq$ 1400 models were selected for further analysis.
The selected models all had the same possible molecular connectivity, but different values for model parameters.
Transcription and translation rates, as well as mRNA and protein degradation terms, were set using physical values from the literature \citep{Milo:2010aa},
and allowed to vary by a scaling factor, see methods.
Model selection was based upon Pareto rank, the prediction and training error across all objectives.
The model population recapitulated key signaling events following TGF-$\beta$ exposure.
We subdivided the response to TGF-$\beta$ exposure into two phases.
First, TGF-$\beta$1/2 signaling initiated a program which downregulated E-cadeherin expression in a MAPK dependent manner while
simultaneously upregulating TGF-$\beta$3 expression. Second, TGF$-\beta$3 secretion initiated an autocrine feedback which upregulated the expression of
mesenchymal markers such as Vimentin and key upstream transcription factors such as LEF-1 in a SMAD dependent manner.
TGF-$\beta$3 expression was also able to sustain $\beta$-catenin release by inhibiting its sequestration by the APC complex through PI3K mediated GSK3,
which was captured by the model (Fig. \ref{fg:F4}B).
Each phase involved the hierarchal expression and/or post-translational modification of several key transcription factors.
During the first phase, stimulation with TGF-$\beta$1/2 (10 a.u.) activated both the SMAD and MAPK pathways.
MAPK activation resulted in the phosphorylation of the transcription factor activator protein 1 (AP-1), which in-turn upregulated the expression of Snail, a well established transcriptional repressor (Fig. \ref{fg:F2}A).
Snail expression was MAPK-dependent; the MEK inhibitor U0126 blocked AP-1 activation and Snail expression following TGF-$\beta$1/2 exposure (Fig. \ref{fg:F2}A, Lane 3).
Similar results were obtained for Slug expression, confirming initial activation through the MAPK pathway (data not shown).
Overexpression of either Snail or Slug upregulated TGF-$\beta$3 expression (Fig. \ref{fg:F2}C) while simultaneously downregulating E-cadeherin expression (Fig. \ref{fg:F2}F).
During the second phase, TGF$-\beta$3 secretion and the subsequent autocrine signaling resulted in the upregulation of mesenchymal marker expression.
The TGF$-\beta$3 induced gene expression program involves a complex hierarchy of transcriptional and post-translational regulatory events.
Absence of E-cadherin indirectly promoted TGF$-\beta$3 expression through the $\beta$-catenin/TCF4 complex following Snail or Slug expression (Fig. \ref{fg:F2}C, Lane 2 or 3).
Conversely, over-expression of E-cadherin inhibited the TGF$-\beta$3 autocrine production by sequestering cytosolic $\beta$-catenin, thereby blocking EMT (Fig. \ref{fg:F2}C, Lane 4 or 5).
TGF$-\beta$3 signaled through the Smad pathway to regulate LEF-1 expression and downstream target EMT genes (Fig. \ref{fg:F2}G).
TGF$-\beta$3 (10 a.u.) in combination with downstream inhibitors (DN-Smad4 and DN-LEF-1) completely inhibited Vimentin expression, while elevating E-cadherin expression (Fig. \ref{fg:F2}H,I).

The predictive power of the ensemble was tested using both cross validation and by comparing simulations with data sets not used for model training.
In whole, 72\% of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
generated from a random starting point. Conversely, we \emph{predicted} approximately 72\% of the training objectives, at a 95\% confidence interval compared to randomized parameters.
The model also captured the temporal gene expression responses of E-cadherin, pSmad2, and LEF-1 (not used for model training) to within one-standard deviation (up to the 48 hr time-point) (Fig. \ref{fg:F2}J-L). Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \citep{Medici:2008fk} that drive the phenotypic conversion.
A listing of objective function values resulting from training, cross validation and the random parameter control is given in the supplement (Fig. \ref{fg:ObjTable}).

\subsection*{Identification of a novel LEF-1 regulator}
During model identification, we found that consistent TGF-$\beta$ induced EMT required an additional regulatory protein.
This protein, which we called hypothetical regulator 1 (YREG1), was required to mediate between SNAIL/SLUG transcriptional activity
and the upregulation of LEF-1 expression following TGF-$\beta$1/2 exposure.
SNAIL/SLUG are well known transcriptional repressors \citep{Hemavathy:2000aa,Hemavathy:2000ab,Dhasarathy:2011aa}, although there are a few studies
which suggest that at least SNAIL can also act as a transcriptional activator \citep{Guaita:2002aa}.
In the model, we assumed the expression of SNAIL/SLUG was likely regulated by AP1/SP1 \citep{Jackstadt:2013aa}.
Thus, upon receiving a TGF-$\beta$1/2 signal, the model predicted enhanced SNAIL/SLUG expression, consistent with experimental observations.
TGF-$\beta$1/2 stimulation also induces LEF-1 expression. However, literature evidence suggested that LEF-1 expression was not strongly dependent upon AP1/SP1 activity \citep{Eastman:1999aa}.
Thus, either SNAIL/SLUG are acting as inducers (contrary to substantial biochemical evidence) or, they are repressing the expression of an intermediate repressor.
Given the biochemical evidence supporting SNAIL/SLUG as repressors, we created YREG1 a hypothetical intermediate repressor whose
expression is downregulated by SNAIL/SLUG. The literature data therefore suggested that YREG1 had two transcriptional targets, LEF-1 and TGF-$\beta$3.
By adding this regulator, our simulations became consistent with training and literature data.
Medici et al. suggested that feedback between $\beta$-catenin and LEF-1 was likely, although this feedback had yet to be identified \citep{Medici:2008fk}.
Low levels of YREG1 expression were present in all simulations to regulate the formation of the $\beta$-catenin-LEF-1 complex.
To test the potency of YREG1, we conducted over-expression and knockdown simulations following the addition of TGF-$\beta$1/2 (Fig. \ref{fg:F4}C and \ref{fg:F4}D).

% TODO: REVISE -----
% In the absence of YREG1, most of the population failed to consistently respond to TGF-$\beta$1/2 exposure compared to the wild-type (Fig. \ref{fig:S8}A).
% Conversely, YREG1 overexpression revealed an exclusively epithelial phenotype following TGF-$\beta$1/2 stimulation (Fig. \ref{fig:S8}B).]
% Overexpression of YREG1 repressed LEF-1 and TGF-$\beta$3 expression,
% thereby not allowing free $\beta$-catenin to form the $\beta$-catenin-LEF-1 complex which promotes mesenchymal gene expression, or SMAD activity following from autocrine TGF-$\beta$3 signaling.
% Taken together, we found that low YREG1 expression was necessary for stabilizing EMT,
% while elevated YREG1 levels limited the extent of EMT induction.

@@

However, the most biologically interesting behavior was exhibited by cells achieving a hybrid phenotype, most notable in a dual treatment condition (black arrow), but also present in a small percentage of untreated cells (Fig. \ref{fg:F3}A).
Models with this hybrid phenotype had elevated Sp1 and NFAT transcriptional activity, resulting in the simultaneous increase of \emph{both} Vimentin and E-cadherin expression (Fig. \ref{fg:F4}A).
Analysis of these hypothetical cells suggested they had \emph{abnormal} signaling; deregulated NFAT expression and nuclear localization promoted E-cadherin expression while TGF$-\beta$1/2 induced Sp1 action promoted Vimentin expression.

% ####Analysis of the connectivity and information flow through the signaling architecture suggested that Sp1 and NFAT action could be manipulated \emph{independently} by simultaneous TGF$-\beta$1/2 and VEGF-A stimulation (Fig. \ref{fg:S1}).

To test this hypothesis, we simulated the response of the network to TGF$-\beta$1/2 and VEGF-A treatment with and without NFATc inhibitors (Fig. \ref{fg:F4}).
As expected, stimulation with VEGF-A (50 a.u.) maintained an epithelial population, while TGF$-\beta$1/2 (10 a.u.) exposure shifted
the population from an epithelial to a mesenchymal phenotype (Fig. \ref{fg:F3}A and Fig. \ref{fg:F3}B).
On the other hand, combined stimulation with TGF$-\beta$1/2 (10 a.u.) and VEGF-A (50 a.u.) increased both E-cadherin and Vimentin expression, resulting in a hybrid phenotype with both epithelial and mesenchymal characteristics (Fig. \ref{fg:F3}C).
Vimentin expression was correlated with high levels of nuclear phosphorylated Sp1, following TGF$-\beta$1/2 exposure.
Conversely, elevated E-cadherin expression depended upon the activity of NFAT transcription factors downstream of VEGF-A stimulation.
To further isolate the role of NFAT on this hybrid state, we simulated the inhibition of NFAT transcriptional activity across all conditions (all else being equal).
NFAT inhibition in combination with VEGF-A and TGF$-\beta$1/2 treatments blocks increased activation of E-cadherin (Fig. \ref{fg:F3}D and Fig. \ref{fg:F3}E).
Lastly, NFATc inhibition in combination with simultaneous TGF$-\beta$1/2 and VEGF-A exposure repressed nearly all E-cadherin expression, shifting nearly the entire population towards a mesenchymal phenotype (Fig. \ref{fg:F4}F). Taken together, high levels of nuclear localized phosphorylated Sp1 correlated with Vimentin expression, while NFATc transcriptional activity was predicted to be critical for maintaining E-cadherin expression in the prescence of competing signals.
In particular, when present, VEGF-A signaling through NFATc was predicted to provide a protective up-regulation of E-cadherin.

@@

\section*{Materials and Methods}
All model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.
